---
title: "Getting started with moultmcmc"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with moultmcmc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: moultmcmc.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

**Note that as of January 2022 the moult package [@erni2013moult] is archived on CRAN because of an incompatibility with the upcoming R version 4.2.x. This vignette therefore relies on a patched version of the moult package for the time being. The patched version can be installed from github using the following command**

```{r setup0, warning=FALSE, message=FALSE}
#install patched moult package from github to avoid R CMD CHECK failures on R-devel
remotes::install_github('pboesu/moult', ref='patch_as_vector_error')
```


```{r setup, warning=FALSE, message=FALSE}
library(moult)
library(moultmcmc)
library(dplyr)
library(ggplot2)
library(rstan)
```

For demonstration purposes we use the `sanderlings` dataset from @underhill1988model. It contains observations of sampling dates `Day` and moult indices `MIndex` for 164 adult Sanderlings (*Calidris alba*) trapped on 11 days in the South Africa in the austral summer of 1978/79. The sample contains 85 pre-moult individuals, 66 in active moult and 13 post-moult individuals. The data are augmented with a column describing categorical moult status `MCat`, so we can demonstrate fitting Type 1 models.

```{r data-preparation, fig.width=7}
data("sanderlings")
sanderlings$MCat <- case_when(sanderlings$MIndex == 0 ~ 1,
                              sanderlings$MIndex == 1 ~3,
                              TRUE ~ 2)
#plot the data
ggplot(sanderlings, aes(y=MIndex, x=Day, col = factor(MCat))) + geom_point() +theme_classic()
```

At present, each Underhill-Zucchini model type is implemented as a standalone function in `moultmcmc`. Function names are composed of the prefix `uz` followed by the numeral for the relevant data type and the postfix `_linpred`. To fit the type 1 model we therefore use the `uz1_linpred()` function. As there are no covariates in the data each linear predictor is specified as an intercept only model using the formula notation `~1`:
```{r uz1-model-mcmc}
uz1 = uz1_linpred(moult_cat_column = "MCat",
            date_column = "Day",
            data = sanderlings,
            start_formula = ~1,
            duration_formula = ~1,
            sigma_formula = ~1,
            init = "auto",
            log_lik = FALSE)
```
A summary table of parameter estimates can be displayed with
```{r uz1-model-summary}
summary_table(uz1)
```
and standard model assessment can be done as for any posterior sample from Stan, by directly accessing the `stanfit` slot of the returned S3 object. For example we can look at the traceplots for the model using `rstan::stan_trace`
```{r uz1-trace-plot, fig.width=7}
stan_trace(uz1$stanfit)
```

For comparison we can fit the same model using maximum likelihood estimation via the `moult` package [@erni2013moult]
```{r uz1-model-ml, fig.width=7}
m1 = moult(MIndex ~ Day,data = sanderlings, type = 1)
summary_table(m1)
compare_plot(m1, uz1, names = c('ML','MCMC'))
```

The comparison plot shows very similar estimates for both methods.

Similarly we can fit the corresponding Type 2 model, that makes use of the full information contained in the moult indices, using the `uz2_linpred()` function. All `moultmcmc` model functions default to an intercept only model, so we do not need to explicitly specify the individual linear predictors here.

```{r uz2-model, fig.width=7}
#fit using MCMC
uz2 = uz2_linpred("MIndex",
                  date_column = "Day",
                  data = sanderlings,
                  log_lik = FALSE)
#fit using ML
m2 = moult(MIndex ~ Day,data = sanderlings, type = 2)
#compare parameter estimates
compare_plot(m2, uz2, names = c('ML','MCMC'))
```

As before the comparison plot shows very similar estimates for both methods. Perhaps more interesting is a comparison of the two model types, which illustrates the gain in precision achieved by exploiting the information contained in the moult indices.
```{r data-type-comparison, fig.width=7}
compare_plot(uz1, uz2, names = c('UZ1','UZ2'))
```

# References

